Thesis: draft

### 2021-05-03 draft.tex
made some small changes in ch.3 and ch. 4

\chapter{Opinion dynamics}
\label{chap3}
The study of opinion dynamics in a group of people, as already mentioned in the introduction, has two major layers: a network layer and an opinion dynamics layer. The network layer, which describes the underlying structure and correlations between the agents/people, is intensively described in the previous chapter, Chapter \ref{chap2}. The current chapter focuses on the second important layer, namely that of the opinion dynamics models. These models need to describe, in a simplified way, how opinions are formed and how they evolve in a group of people that interact in a certain way. Hence, these opinion dynamics are the dynamics that run on the underlying network structure.\\
\newline
Several models have been proposed; here we will review some of the most important ones, with the main focus on the models used in this thesis.\\
\newline
Some major attention will also be given to the content curation/filtering algorithms used by social media companies (see Section \ref{filter}). This is needed because this thesis deals with opinion dynamics on on-line social media, so the aspect of filtering cannot be ignored in the formation and evolution of opinions.\\
\newline
There will also be some attention on the activation mechanism (see Section \ref{actMech}). This is the mechanism that describes the timings in which content/information is exchanged between users \cite{Perra2019}. Such a mechanism may change the network from a static one to a more temporal one, where nodes become active and inactive over time. \\
\newline
There are two major groups of opinion dynamics models, namely binary/discrete models and continuous approaches. These will be discussed in Section \ref{Bin} and Section \ref{Con}, respectively. This thesis only deals with binary opinion dynamics models, so these will be thoroughly described; the continuous models will be quickly reviewed in order to give a more complete overview.\\
\newline
Finally, in Section \ref{stubb}, the introduction of stubborn actors is discussed.

@@

One of the simplest opinion dynamics models is the binary majority model. In this model each agent has one of the two opinions A or B. Initially, these opinions can be equally distributed or not. If the distribution is not equal, but instead eg. 80/20, there is a predominant opinion and a minority opinion. The following updating rule is then imposed on the network: each agent/individual adopts the majority opinion of his/her nearest neighbors \cite{Nguyen2020}. Thus, initially each agent is assigned an opinion; then the opinions are evolved in time: each individual iteratively updates his/her opinion according to the majority opinion of his/her direct neighbors \cite{Nguyen2020}.\\
\newline
Several studies have been performed on this kind of model, searching for answers on several questions such as: what are the conditions that drive the system to converge to unanimity; and, does an opinion, that is initially a majority opinion, remain a majority opinion in the final state \cite{Nguyen2020}? \\
\newline
Most of these studies conclude that, at the end of the opinion evolution, a complete consensus state is obtained where all the agents adopt the same opinion \cite{Nguyen2020}. This is even the case when the opinions are equally and randomly distributed in an Erd\H{o}s-R\'{e}nyi network (with the side note that the degree needs to be high enough), as was shown by Benjamini, Chan, O'Donnell, Tamuz and Tan \cite{Benjamini2016} (also note that even when the opinions are equally distributed, there may still be a slight imbalance due to the random assignement of an opinion to each node; instead of a perfect 50/50 distribution, one may end up with a eg. 49.8/50.2 distribution; in this sense there is still some kind of notion of `minority' and `majority' opinion). It is also found that the rate of convergence increases with increasing edge density and that if the network is sparse enough, consensus will not be reached and the two opinions coexist \cite{Nguyen2020}. This can be explained by the fact that if the network is sparse enough, we may have disconnected parts; each of these disconnected groups may then evolve to a different consensus opinion, so that over the entire network different opinions coexist \cite{Nguyen2020}.

@@

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a more probabilistic/random nature. In this model the fraction of neighbors of user $i$ with opinion A and the fraction of neighbors with opinion B are determined. Then, the opinion of user $i$ is updated: he/she takes on opinion A (or B) with a probability equal to the fraction of his/her neighbors with opinion A (or B) \cite{Perra2019}.\\
\newline
For example, if $70\%$ of the neighbors of user $i$ have an opinion A (and thus $30 \%$ of the neighbors have opinion B), user $i$ will update his/her opinion to opinion A with a probability equal to $0.7$ and to opinion B with a probability equal to $0.3$ \cite{Perra2019}. Thus, user $i$ no longer simply updates his/her opinion according to the majority of opinions carried by his/her neighbors, but the majority opinion still has a higher probability to be taken on by this user.\\
\newline
This model is no longer deterministic, but has a level of randomness, which is an important component of opinion dynamics \cite{Perra2019}.

@@

Another simple binary model, maybe even simpler than the majority model, is the voter model. The model works as follows: each agent in the network has a binary opinion (A or B); at each time step, an agent is randomly selected along with one of its neighbors and, subsequently, the agent adopts the opinion of this neighbor \cite{Castellano2009}. The agent thus mimics its neighbor \cite{Castellano2009}. \\
\newline
The pressure of the majority in only felt in an averaged way, in the sense that a neighbor that carries the majority opinion has more chance to be randomly selected \cite{Castellano2009}.\\
\newline
Lots of studies have been performed on this model. In this thesis, however, we will not make use of the voter model. Hence, we will not go into further detail. Interested readers are referred to references \cite{Castellano2009}, \cite{Fernandez2014} and related references therein.

@@

In bounded confidence models, each node has, beside an opinion $x_i$, a threshold $\epsilon_i$, also called an uncertainty or tolerance (note that $\epsilon$ can be the same for all nodes or can be different for each node) \cite{Castellano2009}\cite{Deffuant2006}. This threshold determines which opinions of neighbors are close enough to the agent's opinion in order to influence that agent; opinions that are beyond this threshold are ignored \cite{Deffuant2006}. In other words, an agent with opinion $x_i$ is only affected/influenced by those neighbors which have an opinion in the interval $\interval{x_i-\epsilon}{x_i+\epsilon}$ \cite{Castellano2009}. \\

@@

The model is based on a compromise strategy: after some discussion and a constructive debate, the opinions of the two agents move closer to each other \cite{Castellano2009}. The final state is a situation where several opinion clusters coexist. This is because the opinions move closer and closer to each other in several clusters; once each cluster is sufficiently far apart from the others (so the difference in opinions between all agents in the two clusters is larger than $\epsilon$), they no longer interact with each other \cite{Castellano2009}. Only agents inside the same cluster keep on interacting with each other, which makes the opinions inside a cluster converge to a single value \cite{Castellano2009}. Usually, the number and sizes of the clusters depends on the threshold $\epsilon$, whereas $\mu$ determines the convergence rate \cite{Castellano2009}.\\

@@

The Hegselmann-Krause model is very similar to the Deffuant model. Again, there are $N$ agents, each with an opinion $x_i$ and a constant threshold $\epsilon$ \cite{Castellano2009}. Each agent interacts with those neighbors with an opinion in the interval $\interval{x_i-\epsilon}{x_i+\epsilon}$ \cite{Castellano2009}. The difference with the Deffuant model lies in the updating rule \cite{Castellano2009}. In the Deffuant model, only the opinion of one randomly selected neighbor is taken into account, whereas the Hegselmann-Krause model considers all compatible neighbors \cite{Castellano2009}. The updating rule becomes \cite{Castellano2009}

@@

The dynamics of this model is similar to the Deffuant model and the final state is also characterized by a number of stationary opinion clusters \cite{Castellano2009}. The number of clusters decreases if $\epsilon$ increased and above a critical value $\epsilon_c$ of the threshold, there can only be one cluster \cite{Castellano2009}.\\

@@

On-line social platforms use algorithms to order and filter the posts that appear on the time-line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies would not use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience a lot more unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that humans can process \cite{Bozdag2013}.\\
\newline
The backside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
\newline
It is important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Perra and Rocha already partly investigated the role of such algorithms on the dynamics of opinions in several networks in one of their previous papers, \textit{``Modelling opinion dynamics in the age of algorithmic personalization''} (see reference \cite{Perra2019}). Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}
\begin{itemize}
	\item Popularity filtering: promoting content that is popular across the platform,
	\item Semantic filtering: promoting posts that are similar to previous consumed posts,
	\item Collaborative filtering: promoting content that people similar or close to us consume.
\end{itemize} 
The following paragraph briefly reviews the filtering algorithms that Perra and Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from his/her friends \cite{Perra2019}. The filtering algorithm then selects $S$ items from this list and shows them on the user's time-line (this is the fundamental idea behind algorithmic personalization: only showing a fraction of the possible posts \cite{Perra2019}); the following selection methods are used \cite{Perra2019}
\begin{itemize}
	\item The reference filtering algorithm (REF): this filtering algorithm randomly selects $S$ items from the list $L_i(t)$.
	\item The old filtering algorithm (OLD): the $S$ oldest post in $L_i(t)$ are displayed on the time-line of the user.
	\item The recent filtering algorithm (REC): shows the $S$ most recent items from $L_i(t)$.
	\item The preference filtering algorithm (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the user and then selects the first $S$ elements. This filtering algorithm is based on the semantic filtering explained above.
\end{itemize}
Perra and Rocha showed that the REF, OLD and REC filtering algorithms behave similar in many cases and that the PR filtering algorithm often displays different behavior (such as a larger formation of echo chambers compared to the other three filtering algorithms) \cite{Perra2019}. In this thesis the REC and REF filtering algorithms are mainly used as a reference/benchmark algorithm and the PR filtering algorithm is used as the main filtering algorithm.

@@

In the most simple opinion dynamics models, which make use of a static network, users are always active and thus always exchanging opinions/posts with each other. In this kind of model the opinion dynamics might go as follows: each time step a random user is selected after which his/her opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all users at the same time during each time step. \\
It might, however, be a more realistic situation if nodes/users become active/inactive over time. This would be a better representation of real social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the users gives the network a more temporal character. The underlying network structure does not change (in the sense that each user/node has a fixed number of neighbors that does not change over time, so no new friends are made or, conversely, no old friends are being defriended), but by making users active and inactive the exchange of information between users becomes time dependent and depends on which users are active and which are not. Hence, in this sense, this does affect the network structure in some way, because if a user, for example, does not become active for a long period of time, he/she does not participate in the exchange of information/opinions and the network structure behaves like if this user/node is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity, clustering coefficient, etc may behave differently. \\
In this thesis we follow the same activation mechanism as was proposed in the paper of Perra and Rocha (see reference \cite{Perra2019}). Every time step users are made active with some probability $p_i$ (thus, every time step, each node becomes active with a probability $p_i$). In this thesis the activation probability is set equal to $p = 0.1$, which is constant for all the nodes. Perra and Rocha chose, in their paper, activation probabilities that are extracted from a power-law ($F(p) \sim p^{-1.5}$, with $p \in \interval{0.01}{1}$) \cite{Perra2019}. They did this based on the analyses of real social networks that shows that the propensity of users to start an on-line social interaction follows a heterogeneous distribution \cite{Perra2012}\cite{Perra2019}. However, they also showed that using a constant activation probability $p$ over the entire population yields results that are qualitatively similar (at least in their case) \cite{Perra2019}. \\
For simplicity, this thesis implements the constant activation probability variant. It might however be an interesting extension to switch to the heterogeneous activation probabilities and investigate the possible impacts on the dynamics. \\
If a user is active, he/she first spreads his current opinion to all his/her neighbors (for example by posting something). Then, the active user updates his own opinion according to the opinions on his/her time-line \cite{Perra2019}. This updating goes according to the implemented opinion dynamics model (eg. majority model, probabilistic majority model,...). In every time step the updating of all active nodes is done simultaneously \cite{Perra2019}.\\
It is important to note that what somebody sees on his/her time-line is the consequence of the interplay between the filtering algorithm and the posted opinions of his/her neighbors. Every user has a ``hidden time-line $L_i(t)$'' which contains the actual time-line of the user at the moment when he/she was active for the last time ($t_{\text{last}}$) plus all the posts of the neighbors from that last time he/she was active till the current time he/she becomes active again ($t$) \cite{Perra2019}. At the time $t$ that the user becomes active again, the filtering algorithm selects $S$ post from this ``hidden time-line $L_i(t)$'' and shows them on the actual time-line of the user; these $S$ items are thus the opinions that the user is exposed to \cite{Perra2019} (see also Section \ref{filter} for more detailed information on the filtering algorithms).

@@

It is well known that in real life, you can find people that are really persistent towards their own opinion and that are not open to opinions of others. They are convinced that they carry the truth with them and that other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. Of course this is put a bit simplistic and there are many shades of gray between this black and white statement. However, there are groups in society, often with more extreme opinions, that are know to be really inflexible towards changing that (extreme) opinion (note that being stubborn and being extreme does not need to go hand in hand, you can obviously be stubborn without being extreme and vice versa; however, it is often widely accepted that there is some kind of correlation/link between people with extreme opinions and being more persistent towards these opinions compared to people with more moderate beliefs; this idea of correlating extreme people with a more stubborn character is also often used in models, see for example \cite{Deffuant2006}\cite{Deffuant2002}\cite{Turner}) (also note that having some resistance to change your opinion to any other opinion does not necessarily need to be a bad thing, it can be seen as a form of critical thinking and as a form of taking things with a grain of salt).\\ 
\newline
Knowing the impact of such stubborn groups on the opinions of others is of great importance, especially with the rise of social media which makes the reach of more extreme and persistent groups a lot bigger. Before the era of social media, one could maybe argue that these extreme and stubborn groups are often isolated and therefore would not reach too many people. So their impact could only be of a minor nature (note that this is only a statement that could have been given, there are of course plenty of examples throughout human history that contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, the rise of communism in the 20th century, etc, all examples of groups that can be considered extreme and that have had a (relatively) wide reach and a significant impact on society). Today, this possible argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no (time and spatial) constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have huge, and largely unknown, impacts. \\
\newline
What is the impact of having stubborn/extreme people in your society? Does it matter if the stubborn people are grouped together or if they are distributed evenly throughout your system? Are there ways to reduce the possible impacts of stubbornness and/or extremism? Can these stubborn groups lead to polarization or even overtake the opinions in the system? Or is introducing some kind of stubbornness, on the other hand, a protective factor towards extremism/polarization? To answer these questions, several models have been proposed to include some kind of stubbornness/resistance in the system under study.

@@

The bounded confidence models, discussed in Section \ref{bounded}, can be easily extended to included some notion of stubbornness. Guillaume Deffuant proposed such an extension where stubbornness is correlated to extremism \cite{Deffuant2006}. The model he proposed goes as follows. Remember that in the bounded confidence models, the agents have an opinion $x_i$ (which is a real number, often between $\interval{0}{1}$ or between $\interval{-1}{1}$) and a threshold $\epsilon_i$, which determines the range of opinions that can influence agent $i$ (agents with opinions between $\interval{x_i - \epsilon_i}{x_i + \epsilon_i}$ are capable of influencing agent $i$). Guillaume Deffuant then introduced a fraction of extremists, $f_e$, which can have either the extreme opinion $-1$ or the extreme opinion $1$ \cite{Deffuant2006}. These extremists are given a threshold equal to $\epsilon_e$ \cite{Deffuant2006}. The other agents are considered to be moderate and can have opinions between $-1$ and $1$. The moderates are given a threshold $\epsilon$, with $\epsilon > \epsilon_e$ \cite{Deffuant2006}. This last condition implicitly denotes the resistance of the extremes towards other opinions in contrast to the more flexible nature of the moderates, which is again based on the idea that people with extreme opinions will be more convinced of that opinion and will be less likely to change it.\\
\newline
Since this thesis is not concerned with bounded confidence models, we will not discuss the results of this model here. Interested readers can find them in reference \cite{Deffuant2006}. \\
\newline
Another example of introducing some kind of persistence towards the own opinion of the nodes is the ``hardening of positions'' that was briefly discussed at the end of Section \ref{classMod}.

@@

Playing around with the threshold $\epsilon_i$ of the agents to include a notion of stubbornness, such as one can do for the bounded confidence models, is obviously not possible for binary/discrete models. There are, however, many other ways to include some kind of stubborn actors in these models. In the following, two different possibilities will be discussed. \\

@@

One small remark must be made here: previously, the link between stubbornness and extremism was often made and it was often assumed that the more extreme somebodies opinions are, the more persistent that somebody is towards those opinions. In binary models, however, there are only two possible opinions that can be taken on and in that sense it is less useful to talk about extreme and moderate opinions. However, we can still have people that rather keep their own opinion than be convinced by friends to switch to the other opinion. Whether this resistance is a sign of critical thinking or ignorance/being convinced that you know it best, is up to the reader to decide. The models simply introduce some type of resistance and make no distinction between possible underlying reasons for this resistance.\\

@@

One way of creating stubborn actors in the binary model is by giving each node $i$ a stubbornness/resistance parameter $r_i$. This resistance parameter can have values between 0 and 1, $0 \leqslant r_i \leqslant 1$. 
	\item If $r_i = 0$, the node is not stubborn and can be easily convinced of the value of the other opinion (according to which opinion dynamics model that is implemented: majority, probabilistic majority,...).
	\item If $r_i = 1$, the node is completely stubborn and there is no chance of convincing him/here of the other idea.

>>>>

\chapter{Opinion dynamics}
\label{chap3}
The study of opinion dynamics in a group of people, as already mentioned in the introduction, has two major layers: a network layer and an opinion dynamics layer. The network layer, which describes the underlying structure and correlations between the agents/people, is intensively described in the previous chapter, Chapter \ref{chap2}. The current chapter focuses on the second important layer, that of the opinion dynamics models. These models need to describe, in a simplified way, how opinions are formed and how they evolve in a group of people that interact in a certain way. Hence, these opinion dynamics are the dynamics that run on the underlying network structure.\\
\newline
Several models have been proposed; here we will review some of the most important ones, with the main focus on the models used in this thesis.\\
\newline
Some major attention will also be given to the content curation/filtering algorithms used by social media companies (Section \ref{filter}). This is needed because this thesis deals with opinion dynamics on on-line social media, so the aspect of filtering cannot be ignored in the formation and evolution of opinions.\\
\newline
There will also be some attention on the activation mechanism (Section \ref{actMech}). This is the mechanism that describes the timings in which content/information is exchanged between users \cite{Perra2019}. Such a mechanism changes the network from a static one to a more temporal one, where nodes become active and inactive over time. \\
\newline
There are two major groups of opinion dynamics models, namely binary/discrete models and continuous approaches. These will be discussed in Section \ref{Bin} and Section \ref{Con}, respectively. This thesis only deals with binary opinion dynamics models, so these will be thoroughly described; the continuous models will be quickly reviewed in order to give a more complete overview.\\
\newline
Finally, in Section \ref{stubb}, the introduction of stubborn actors is discussed.

@@

One of the simplest opinion dynamics models is the binary majority model. In this model each agent has one of the two opinions A or B. Initially, these opinions can be equally distributed or not. If the distribution is not equal, but instead eg. 80/20, there is a predominant opinion and a minority opinion. The following updating rule is then imposed on the network: each agent/individual adopts the majority opinion of his/her nearest neighbors \cite{Nguyen2020}. Initially each agent is assigned an opinion; then the opinions are evolved in time: each individual iteratively updates his/her opinion according to the majority opinion of his/her direct neighbors \cite{Nguyen2020}.\\
\newline
Several studies have been performed on this kind of model, searching for answers on several questions such as: What are the conditions that drive the system to converge to unanimity; and, Does an opinion, that is initially a majority opinion, remain a majority opinion in the final state \cite{Nguyen2020}? \\
\newline
Most of these studies conclude that, at the end of the opinion evolution, a complete consensus state is obtained where all the agents adopt the same opinion \cite{Nguyen2020}. This is even the case when the opinions are equally and randomly distributed in an Erd\H{o}s-R\'{e}nyi network (with the side note that the degree needs to be high enough), as was shown by Benjamini, Chan, O'Donnell, Tamuz and Tan \cite{Benjamini2016}. It is also found that the rate of convergence increases with increasing edge density and that if the network is sparse enough, consensus will not be reached and the two opinions coexist \cite{Nguyen2020}. This can be explained by the fact that if the network is sparse enough, we may have disconnected parts; each of these disconnected groups may then evolve to a different consensus opinion, so that over the entire network different opinions coexist \cite{Nguyen2020}.

@@

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a more probabilistic/random nature. In this model the fraction of neighbors of user $i$ with opinion A, $p_\text{A}$, and the fraction of neighbors with opinion B, $p_\text{B}$, are determined. Then, the opinion of user $i$ is updated: he/she takes on opinion A (or B) with a probability equal to $p_\text{A}$ (or $p_\text{B}$) \cite{Perra2019}.\\
\newline
For example, if $70\%$ of the neighbors of user $i$ have an opinion A (and thus $30 \%$ of the neighbors have opinion B), user $i$ will update his/her opinion to opinion A with a probability equal to $0.7$ and to opinion B with a probability equal to $0.3$ \cite{Perra2019}. Thus, user $i$ no longer simply updates his/her opinion according to the majority of opinions carried by his/her neighbors, but the majority opinion still has a higher probability to be taken on by this user.\\
\newline
This model is no longer deterministic, but has a level of randomness, which is an important component of opinion dynamics \cite{Perra2019}.

@@

Another simple binary model is the voter model. The model works as follows: each agent in the network has a binary opinion (A or B); at each time step, an agent is randomly selected along with one of its neighbors and, subsequently, the agent adopts the opinion of this neighbor \cite{Castellano2009}. The agent thus mimics its neighbor \cite{Castellano2009}. \\
\newline
The pressure of the majority in only felt in an averaged way, in the sense that a neighbor that carries the majority opinion has more chance to be randomly selected \cite{Castellano2009}.\\
\newline
Lots of studies have been performed on this model. In this thesis, however, we will not make use of the voter model. Hence, we will not go into further detail. Interested readers are referred to references \cite{Castellano2009}, \cite{Fernandez2014} and related references therein.

@@

In bounded confidence models, each node has, beside an opinion $x_i$, a threshold $\epsilon_i$, also called an uncertainty or tolerance ($\epsilon$ can be the same for all nodes or can be different for each node) \cite{Castellano2009}\cite{Deffuant2006}. This threshold determines which opinions of neighbors are close enough to the agent's opinion in order to influence that agent; opinions that are beyond this threshold are ignored \cite{Deffuant2006}. In other words, an agent with opinion $x_i$ is only affected/influenced by those neighbors which have an opinion in the interval $\interval{x_i-\epsilon}{x_i+\epsilon}$ \cite{Castellano2009}. \\

@@

The model is based on a compromise strategy: after some discussion and a constructive debate, the opinions of the two agents move closer to each other \cite{Castellano2009}. The final state is a situation where several opinion clusters coexist. This is because the opinions move closer and closer to each other in several clusters; once each cluster is sufficiently far apart from the others (the difference in opinions between all agents in the two clusters is larger than $\epsilon$), they no longer interact with each other \cite{Castellano2009}. Only agents inside the same cluster keep on interacting with each other, which makes the opinions inside a cluster converge to a single value \cite{Castellano2009}. Usually, the number and sizes of the clusters depends on the threshold $\epsilon$, whereas $\mu$ determines the convergence rate \cite{Castellano2009}.\\

@@

The Hegselmann-Krause model is similar to the Deffuant model. Again, there are $N$ agents, each with an opinion $x_i$ and a constant threshold $\epsilon$ \cite{Castellano2009}. Each agent interacts with those neighbors with an opinion in the interval $\interval{x_i-\epsilon}{x_i+\epsilon}$ \cite{Castellano2009}. The difference with the Deffuant model lies in the updating rule \cite{Castellano2009}. In the Deffuant model, only the opinion of one randomly selected neighbor is taken into account, whereas the Hegselmann-Krause model considers all compatible neighbors \cite{Castellano2009}. The updating rule becomes \cite{Castellano2009}

@@

The dynamics of this model is similar to the Deffuant model and the final state is also characterized by a number of stationary opinion clusters \cite{Castellano2009}. The number of clusters decreases if $\epsilon$ increases and above a critical value $\epsilon_c$ of the threshold, there can only be one cluster \cite{Castellano2009}.\\

@@

On-line social platforms use algorithms to order and filter the posts that appear on the time-line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies would not use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that humans can process \cite{Bozdag2013}.\\
\newline
The downside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
\newline
It is important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Perra and Rocha already partly investigated the role of such algorithms on the dynamics of opinions in several networks in one of their previous papers, \textit{``Modelling opinion dynamics in the age of algorithmic personalization''} \cite{Perra2019}. Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}: (i) Popularity filtering: promoting content that is popular across the platform; (ii) Semantic filtering: promoting posts that are similar to previous consumed posts; (iii) Collaborative filtering: promoting content that people similar or close to us consume.\\ 
\newline
The following paragraph briefly reviews the filtering algorithms that Perra and Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from his/her friends \cite{Perra2019}. The filtering algorithm then selects $S$ items from this list and shows them on the user's time-line (this is the fundamental idea behind algorithmic personalization: only showing a fraction of the possible posts \cite{Perra2019}); the following selection methods are used \cite{Perra2019}: (i) The reference filtering algorithm (REF): this filtering algorithm randomly selects $S$ items from the list $L_i(t)$; (ii) The old filtering algorithm (OLD): the $S$ oldest post in $L_i(t)$ are displayed on the time-line of the user; (iii) The recent filtering algorithm (REC): shows the $S$ most recent items from $L_i(t)$; (iv) The preference filtering algorithm (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the user and then selects the first $S$ elements. This filtering algorithm is based on the semantic filtering explained above.\\
\newline
Perra and Rocha showed that the REF, OLD and REC filtering algorithms behave similar in many cases and that the PR filtering algorithm often displays different behavior (such as a larger formation of echo chambers compared to the other three filtering algorithms) \cite{Perra2019}. In this thesis the REC and REF filtering algorithms are mainly used as a reference/benchmark algorithm and the PR filtering algorithm is used as the main filtering algorithm.

@@

Most of the simple opinion dynamics models make use of a static network in which users are always active and always exchanging opinions/posts with each other. In this kind of model the opinion dynamics might go as follows: each time step a random user is selected after which his/her opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all users at the same time during each time step. \\
\newline
It might, however, be a more realistic situation if nodes/users become active/inactive over time. This would be a better representation of real social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the users gives the network a more temporal character. The underlying network structure does not change (in the sense that each user/node has a fixed number of neighbors that does not change over time, so no new friends are made or, conversely, no old friends are being defriended), but by making users active and inactive the exchange of information between users becomes time dependent and depends on which users are active and which are not. Hence, this does affect the network structure in some way, because if a user, for example, does not become active for a long period of time, he/she does not participate in the exchange of information/opinions and the network structure behaves like if this user/node is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity, clustering coefficient, etc may behave differently. \\
\newline
In this thesis we follow the same activation mechanism as was proposed in the paper of Perra and Rocha \cite{Perra2019}. Every time step users are made active with some probability $p_i$ (thus, every time step, each node becomes active with a probability $p_i$). In this thesis the activation probability is set equal to $p = 0.1$, which is constant for all the nodes. Perra and Rocha chose, in their paper, activation probabilities that are extracted from a power-law ($F(p) \sim p^{-1.5}$, with $p \in \interval{0.01}{1}$) \cite{Perra2019}. They did this based on the analyses of real social networks that shows that the propensity of users to start an on-line social interaction follows a heterogeneous distribution \cite{Perra2012}\cite{Perra2019}. However, they also showed that using a constant activation probability $p$ over the entire population yields results that are qualitatively similar (at least in their case) \cite{Perra2019}. \\
\newline
For simplicity, this thesis implements the constant activation probability variant. It might however be an interesting extension to switch to the heterogeneous activation probabilities and investigate the possible impacts on the dynamics. \\
\newline
If a user is active, he/she first spreads his current opinion to all his/her neighbors. Then, the active user updates his own opinion according to the opinions on his/her time-line \cite{Perra2019}. This updating goes according to the implemented opinion dynamics model (eg. majority model, probabilistic majority model,...). In every time step the updating of all active nodes is done simultaneously \cite{Perra2019}.

@@

It is well known that in real life, you can find people that are persistent towards their own opinion and that are not open to opinions of others. They are convinced that they carry the truth with them and that other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. Of course this is put a bit simplistic and there are many shades of gray between this black and white statement. However, there are groups in society, often with more extreme opinions, that are know to be really inflexible towards changing that (extreme) opinion (note that being stubborn and being extreme does not need to go hand in hand, you can be stubborn without being extreme and vice versa; however, it is often widely accepted that there is some kind of correlation/link between people with extreme opinions and being more persistent towards these opinions compared to people with more moderate beliefs; this idea of correlating extreme people with a more stubborn character is also often used in models, see for example \cite{Deffuant2006}\cite{Deffuant2002}\cite{Turner}) (also note that having some resistance to change your opinion to any other opinion does not necessarily need to be a bad thing, it can be seen as a form of critical thinking and as a form of taking things with a grain of salt).\\ 
\newline
Knowing the impact of stubborn groups on the opinions of others is of great importance, especially with the rise of social media which makes the reach of more extreme and persistent groups a lot bigger. Before the era of social media, one could maybe argue that these extreme and stubborn groups are often isolated and therefore would not reach too many people. So their impact could only be of a minor nature (this is only a statement that could have been given, there are of course plenty of examples throughout human history that may contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, the rise of communism in the 20th century, etc, all examples of groups that can be considered extreme and that have had a (relatively) wide reach and a significant impact on society). Today, this possible argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no (time and spatial) constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have huge, and largely unknown, impacts. \\
\newline
What is the impact of having stubborn/extreme people in your society? Does it matter if the stubborn people are grouped together or if they are distributed evenly throughout your system? Are there ways to reduce the possible impacts of stubbornness and/or extremism? Can these stubborn groups lead to polarization or even overtake the opinions in the system? Or is introducing some kind of stubbornness, on the other hand, a protective factor towards extremism/polarization? To answer these questions, several models have been proposed to include some kind of stubbornness/resistance in the system under study.

@@

The bounded confidence models, discussed in Section \ref{bounded}, can be easily extended to included some notion of stubbornness. Guillaume Deffuant proposed such an extension where stubbornness is correlated to extremism \cite{Deffuant2006}. The model he proposed goes as follows. Remember that in the bounded confidence models, the agents have an opinion $x_i$ (which is a real number between $\interval{-1}{1}$) and a threshold $\epsilon_i$, which determines the range of opinions that can influence agent $i$ (agents with opinions between $\interval{x_i - \epsilon_i}{x_i + \epsilon_i}$ are capable of influencing agent $i$). Guillaume Deffuant then introduced a fraction of extremists, $f_e$, which can have either the extreme opinion $-1$ or the extreme opinion $1$ \cite{Deffuant2006}. These extremists are given a threshold equal to $\epsilon_e$ \cite{Deffuant2006}. The other agents are considered to be moderate and can have opinions between $-1$ and $1$. The moderates are given a threshold $\epsilon$, with $\epsilon > \epsilon_e$ \cite{Deffuant2006}. This last condition implicitly denotes the resistance of the extremes towards other opinions in contrast to the more flexible nature of the moderates, which is based on the idea that people with extreme opinions will be more convinced of that opinion and will be less likely to change it.\\
\newline
Since this thesis is not concerned with bounded confidence models, we will not discuss the results of this model here. Interested readers can find them in reference \cite{Deffuant2006}. \\
\newline
Another example of introducing some kind of persistence towards the own opinion of the nodes is the ``hardening of positions'' that was briefly discussed at the end of Section \ref{classMod}.

@@

Playing around with the threshold $\epsilon_i$ of the agents to include a notion of stubbornness, such as one can do for the bounded confidence models, is not possible for binary/discrete models. There are, however, many other ways to include some kind of stubborn actors in these models. In the following, two different possibilities will be discussed. \\

@@

One small remark must be made here: previously, the link between stubbornness and extremism was often made and it was often assumed that the more extreme somebodies opinions are, the more persistent that somebody is towards those opinions. In binary models, however, there are only two possible opinions that can be taken on and in that sense it is less useful to talk about extreme and moderate opinions. However, we can still have people that rather keep their own opinion than be convinced by friends to switch to the other opinion.\\

@@

One way of creating stubborn actors in the binary model is by giving each node $i$ a stubbornness/resistance parameter $r_i$. This resistance parameter can have values between 0 and 1. If $r_i = 0$, the node is not stubborn and can be easily convinced of the value of the other opinion (according to which opinion dynamics model that is implemented: majority, probabilistic majority,...). If $r_i = 1$, the node is completely stubborn and there is no chance of convincing him/here of the other idea. If $0 < r_i < 1$, the node has some resistance but can still be convinced to switch to the other opinion, even though he/she will be less likely to do so. An uniform random number $f$ between 0 and 1 is drawn. If $f > r_i$, the node will update his/her opinion; if $f \leqslant r_i$, the node is not convinced and keeps his/her own opinion.\\
